Recent demography drives changes in linked selection across the maize genome

### 2015-10-28 manuscript.tex
intro updated

\title{Demography and linked selection in wild and domesticated maize}
\author{Timothy M. Beissinger\affil{1}{Dept. of Plant Sciences, University of California, Davis, CA, USA}\affil{2}{US Department of Agriculture, Agricultural Research Service, Columbia, MO, USA}\affil{3}{Division of Plant Sciences, University of Missouri, Columbia, MO, USA}, Li Wang\affil{4}{Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, IA, USA}\affil{5}{Genome Informatics Facility, Iowa State University, Ames, IA, USA}, Kate Crosby\affil{1}{}, Arun Durvasula\affil{1}{}, Matthew B. Hufford\affil{4}{}, \and Jeffrey Ross-Ibarra\affil{1}{}\affil{6}{Genome Center and Center for Population Biology, University of
    California, Davis, CA, USA} }

@@

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to changes in genetic variabiligy \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. Importantly, demography and selection do not operate independently to determine genetic variability \cite{li2012}. Instead, each of these forces is constantly masking and obfuscating the other, ultimately meaning that the selective history of a species cannot be properly evaluated without a thorough estimate of demographic history, and demography cannot be evaluated without an understanding of selection.

For instance, to estimate the historical demographic parameters of a population, researchers often utilize supposed neutrally-evolving DNA  \cite{luikart2003, gutenkunst2009}. However, it has been called into question whether reasonable approaches exist
for identifying neutral regions of the genome, since the effects of
selection can be wide-ranging \cite{li2012, slotte2014}. In \emph{Drosophila}, for example, the majority of the genome is
impacted by the effects of selection through linkage
\cite{sella2009}. One technique that has been utilized to estimate demographic parameters independently of selective parameters is to limit analyses to only sites far from genes
\cite{gazave2014}, but as \emph{Drosophila} demonstrates, it is
difficult to be certain, without species-specific knowledge, that sites even in gene-poor regions of the
genome are not influenced by linked selection. Ultimately, an understanding
of which sites have the potential to be adaptive, and how these affect
genome-wide patterns of variability through linkage, is required for
reasonable demographic inference. At the same time, since the effects of demography alter genetic variability genome-wide \cite{akey2009}, and because the efficiency of selection depends on demographic characteristics such as population size \cite{kimura1963}, selection is equally difficult to quantify without properly accounting for demographic history.

Maize (\emph{Zea mays}) represents an excellent organism to study this interplay. Maize is a species of tremendous importance worldwide as both a staple crop \cite{shiferaw2011} and as a model for understanding plant evolution \cite{strable2009}. 
Archaeological and genetic studies have
established that maize domestication is likely to have taken place in
 Mexico approximately 9,000 years bp
\cite{smith1995,matsuoka2002}. Teosinte, the most
recent wild ancestor to maize, remains extant throughout much of the
Americas \cite{wilkes1967}. Additionally, several large-effect
domestication loci \cite{doebley1995, wills2013, wang2015} and putative domestication
regions \cite{hufford2012} have been identified. But despite all that is
known about maize domestication, the demographic parameters of the
domestication process remain uncertain. Specifically, the size of the
maize domestication bottleneck has not been estimated independently of
the bottleneck's duration, nor are there sequence-based estimates of the effective
population size of modern maize. At the same time, the predominant form or forms of selection that have taken place during and since domestication remain unknown. Sequence information from maize and teosinte may therefore be utilized to address these questions, and to better explain the interaction between selection and demography in a system with complex but quantifiable selective and demographic histories.

To that end, the objectives of our study were to 1) investigate the
relative importance of different forms of selection on whole-genome
variability, using maize and teosinte as a model; 2) research the impact that
demographic history has had on genetic variability in maize, and how
this compares to the impact of a different demographic history in teosinte; and 3) precisely estimate the parameters of the
maize domestication bottleneck. Gaining an improved understanding of the interaction between demography and selection in maize will help to increase the understanding of these phenomena across a wide array of organisms.

%% \dropcap{D}omesticated plant species evolve in a unique fashion
%% compared to their wild counterparts \cite{doebley2006}.
%% This is a result of both the anthropomorphic nature of artificial selection on
%% domesticates \cite{purugganan2009} as well as the demographic characteristics of the domestication
%% bottleneck(s) that they tend to have experiencedtu
%% \cite{ross2007}. However, the
%% complex interplay between selective pressures and demographic
%% limitations, and the impact that this interplay has on identifying
%% selection and understanding demography, is not fully understood. Although a large body of
%% research that involves searching the genomes of domesticated species for evidence
%% of positive selection exists \cite{hufford2012, he2011, vigouroux2002, chapman2008}, these studies tend to focus on
%% identifying or mapping particular genes or regions that play an
%% important role in phenotypic evolution. In contrast, knowledge regarding the impacts that demography and
%% selection have on whole-genome patterns of genetic variability remains limited.

\section{Results}
%\subsection{Patterns of diversity differ between genic and  intergenic regions of the genome} %JRI
To investigate how demography and linked selection have shaped patterns of diversity in maize and teosinte, we analyzed data from 23 maize and 13 teosinte genomes from the maize HapMap 2 and HapMap 3 projects \cite{chia2012, hapmap3}.
We find broad differences in genic and intergenic diversity consistent with earlier results  \cite{hufford2012}(Figure \ref{fig:diversity}).  In maize, mean pairwise diversity ($\pi$) within genes was significantly lower than at sites at least 5 kb away from genes (0.00668 vs 0.00691, $p<2\times 10^{-44}$). 
Diversity differences in teosinte are even more pronounced (0.0088 vs. 0.0115, $p\approx 0$). 
Differences were also apparent in the site frequency spectrum, with mean Tajima's D positive in genic regions in both maize (0.4) and teosinte (0.013) but negative outside of genes (-0.087 in maize and -0.25 in teosinte, $p\approx 0$ for both comparisons).
These observations suggest that diversity in genes is not evolving neutrally, but instead is reduced by the impacts of selection on linked sites.

@@

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation and/or non-coding variants as primary sources of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  \tmb{Jeff, rewrote this part, and I like your writing here, but as written I don't think we need to mention $s_b?$}
Assuming a mutation rate of $3 \times 10^{-8}$ \cite{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \cite{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \cite[e.g.]{studer2011}.
Indeed, many adaptive traits in both maize \cite{wallace2014} and teosinte \cite{weber2008} are highly quantitative,  and adaptation in both maize \cite{hufford2012} and teosinte \cite{pyhajarvi2013complex} has involved selection on regulatory variation.

>>>>

\title{Recent demography drives changes in linked selection across the maize genome}
\author{Timothy M. Beissinger\affil{1}{US Department of Agriculture, Agricultural Research Service, Columbia, MO, USA}\affil{2}{Division of Plant Sciences, University of Missouri, Columbia, MO, USA}\affil{3}{Dept. of Plant Sciences, University of California, Davis, CA, USA}, Li Wang\affil{4}{Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, IA, USA}\affil{5}{Genome Informatics Facility, Iowa State University, Ames, IA, USA}, Kate Crosby\affil{1}{}, Arun Durvasula\affil{1}{}, Matthew B. Hufford\affil{4}{}, \and Jeffrey Ross-Ibarra\affil{1}{}\affil{6}{Genome Center and Center for Population Biology, University of
    California, Davis, CA, USA} }

@@

\dropcap{T}he genetic properties of populations are determined by a constant interplay between genetic drift and selection. Drift is the consequence of a finite population size, such that random sampling of gametes every generation will incrementally lead to changes in genetic variabiligy \cite{dobzhansky1957}. In other words, the demographic history of a population, which includes its population size fluctuations and mating patterns, has a long-lasting impact on the patterns of DNA variability displayed by that population. Conversely, selection systematically alters a population by favoring particular alleles at the expense of others. In contrast to drift, which results from demographic history, selection is the result of the relationship between genotypes, phenoytpes, and ultimately fitness. A classical approach for studying selection is to remove the genome-wide effects of demography and assume that selection will leave locus-specific evidence \cite{Wall2002, akey2009}. Conversely, demography is often researched by excluding potentially selected sites \cite{voight2005, luikart2003, gutenkunst2009}.

However, recent work has shown that demography and selection do not operate independently to determine genetic variability, in large part because linkage allows the effects of selection to be wide-ranging \cite{li2012,slotte2014}. For example, in \emph{Drosophila} the majority of the genome is
impacted by the effects of selection through linkage
\cite{sella2009}. One technique that has been utilized to estimate demographic parameters independently of selective parameters is to limit analyses to only sites far from genes\cite{gazave2014}, but as \emph{Drosophila} demonstrates, it is
difficult to be certain, without species-specific knowledge, that sites even in gene-poor regions of the genome are not influenced by linked selection. Therefore, characterizing selection, demography or both, first requires an understanding of linked selection. It has already been established that patterns of linked selection are heavily influenced by population size ($N_e$) \cite{cutter2013, slotte2014}, but nearly all species including humans \cite{reich1998}, domesticates \cite{hyten2006, bovine2009}, ecological \cite{storfer2014}, and model species \cite{duchen2013} have experienced recent or ancient changes in $N_e$. Although much is known about how the long-term average $N_e$ affects linked selection, much less is understood about the immediate effects and dynamics of changing in $N_e$ on patterns of linked selection.

Because of its relatively simple demographic history and abundant availability of genomic tools, maize (\emph{Zea mays}) represents an excellent organism to study this phenomenon. Maize is a species of tremendous importance worldwide as both a model species \cite{strable2009} and a staple crop \cite{shiferaw2011}. Archaeological and genetic studies have established that maize domestication began in
Central Mexico at least 9,000 years bp
\cite{smith1995,matsuoka2002}, and involved a population bottleneck followed by recent expansion \cite{wright2005}. Because of this simple but dynamic demographic history, patterns of linked selection in wild and domesticated maize can be used to understand the effects of changing $N_e$ on linked selection. Teosinte, which remains extant throughout much of the Americas \cite{wilkes1967}, is the wild ancestor to maize. Additionally, several large-effect
domestication loci \cite{doebley1995, wills2013, wang2015} and putative domestication
regions \cite{hufford2012} have been identified in maize. But despite all that is
known about maize domestication, the demographic parameters of the
domestication process remain uncertain. Specifically, the size of the
maize domestication bottleneck has not been estimated independently of
the bottleneck's duration, nor are there sequence-based estimates of the effective
population size of modern maize. At the same time, the predominant form or forms of selection that have taken place during and since domestication, and how linkage impacts the evidence and patterns of selection across the genome, remain unknown.

For this study, we begin by estimating the parameters of maize domestication based on whole-genome sequencing data. Next, we utilized these estimates to 1) investigate the relative importance of different forms of selection on genetic
variability in the recent and acient past; and 2) research the impact that demographic history, specifically changes in $N_e$, has had on linked selection in maize, and how this compares to the impact of a different demographic history in teosinte. Gaining an improved understanding of the impacts of a dynamic demography on whole-genome variability in maize will help to increase the understanding of these phenomena across all organisms, since changes in $N_e$ are pervasive.

\section{Results}

\subsection{Patterns of diversity differ between genic and  intergenic regions of the genome} %JRI
To investigate how demography and linked selection have shaped patterns of diversity in maize and teosinte, we analyzed data from 23 maize and 13 teosinte genomes from the maize HapMap 2 and HapMap 3 projects \cite{chia2012, hapmap3}.
We find broad differences in genic and intergenic diversity consistent with earlier results  \cite{hufford2012}(Figure \ref{fig:diversity}).  In maize, mean pairwise diversity ($\pi$) within genes was significantly lower than at sites at least 5 kb away from genes (0.00668 vs 0.00691, $p<2\times 10^{-44}$). 
Diversity differences in teosinte are even more pronounced (0.0088 vs. 0.0115, $p\approx 0$). 
Differences were also apparent in the site frequency spectrum, with mean Tajima's D positive in genic regions in both maize (0.4) and teosinte (0.013) but negative outside of genes (-0.087 in maize and -0.25 in teosinte, $p\approx 0$ for both comparisons).
These observations suggest that diversity in genes is not evolving neutrally, but instead is reduced by the impacts of selection on linked sites.

@@

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation and/or non-coding variants as primary sources of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}. Assuming a mutation rate of $3 \times 10^{-8}$ \cite{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \cite{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \cite[e.g.]{studer2011}.
Indeed, many adaptive traits in both maize \cite{wallace2014} and teosinte \cite{weber2008} are highly quantitative,  and adaptation in both maize \cite{hufford2012} and teosinte \cite{pyhajarvi2013complex} has involved selection on regulatory variation.
